---
title: "Analysis of Vocabulary and Education"
author: "Feray Ece Topcu - BDA 503 - Fall 2017"
output:
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Vocabulary and Education Dataset

  [Vocabulary and Education dataset](https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/car/Vocab.csv) supplied by U.S. General Social Surveys, 1972-2004. It contains vocabulary test score of respondents according to some parameters in U.S. between 1974-2004.

###Content
  The Vocab dataset has 21,638 rows and 5 columns.

+ ***year:*** Year of the survey.
+ ***sex:*** Sex of the respondent, Female or Male.
+ ***education:*** Education in years.
+ ***vocabulary:*** Vocabulary test score: number correct on a 10-word test.

*************

###Initial Exploratory Analysis

This dataset includes 21,638 observations  and 5 variables as listed below. I try to find correlation between Vocabulary test grade and Education. Additionally, I examine the difference between two genders according the years.

```{r,echo=FALSE,results="hide",include=FALSE}
urlfile<-'https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/car/Vocab.csv'
dsin<-read.csv(url(urlfile))
library(dplyr)#glimpse function
library(ggplot2)# graphs and other staff.
```

```{r}
#Quick review of dataset
glimpse(dsin)
```

```{r}
#Which years are included on research and average education in year and vocabulary test score according to years.

g3<-dsin %>% group_by(year) %>% summarise(Avg_Educ_In_Year = round(sum(education)/n(),2) ,Avg_Vocab_Test=round(sum(vocabulary)/n(),2))
g3

ggplot(g3, aes(year)) + 
  geom_point(aes(y = Avg_Vocab_Test, colour = "Average Vocabulary Test Score")) +geom_point(aes(y = Avg_Educ_In_Year, colour = "Average Education in Year"))+scale_colour_manual(values=c("red", "blue")) +theme_minimal() + xlab("Year") + ylab("Average") +ggtitle("Is there any correlation?")

##Expand the output with genders.
a<-dsin %>% group_by(year,sex)%>%summarise(cnt=n(),edu_sum=sum(education),voc_sum=sum(vocabulary),avg_edu=round(edu_sum/cnt,2), avg_vocab=round(voc_sum/cnt,2))
##Female
a%>%filter(sex=="Female")
##Male
a%>%filter(sex=="Male")
##Draw a graph which explains the difference between genders according to year.
ggplot(data=a, aes(x=sex,y=year,color=avg_vocab,size=5))+ coord_flip() + geom_point() + scale_color_continuous(low="red",high="black",guide="colourbar") +ggtitle("Who failed?") +
xlab("Gender") + ylab("Years")
```

*************

##References

##Dataset 
  [https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/car/Vocab.csv]()
  
##Analysis
  [https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html](Dplyr)
  
  [https://www.rstudio.com/resources/cheatsheets/]()
  
  [https://mef-bda503.github.io/files/02_Tidyverse.html#1](Class Presentation about Tidyverse & Dplyr) 
  

##Graphs
  [https://plot.ly/ggplot2/](Thanks God! I found it!)
  
  [http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html](ggplot2)
  
  [http://ggplot2.tidyverse.org/](ggplot2)
  
  [https://stackoverflow.com/questions/3777174/plotting-two-variables-as-lines-using-ggplot2-on-the-same-graph](Special thanks to Stackoverflow, of course! )
  

  
  
  